Leveraging genomics to detect trait selection from within and between species variations

### 2023-07-25 manuscript/main.tex
Updated manuscript

\title{Trait selection from within and between species variations while integrating genomics}
\rhead{\scshape Trait selection from within and between species variations}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
    Conversely, using between species variation, current methods usually discards within species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

@@

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.

First, leveraging individuals variation within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given loci), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
Gene expression exhibit such accumulation, and divergence in expression accumulates faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for a selection of trait in a pair of species~\cite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
Secondly and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Altogether, methods at the phylogenetic scale easily mis-classify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that is does not requires estimate of population size (within species) nor of time of speciation (between species).
For population-geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For a phylogeneticist, our study can be seen as a extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

@@

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutation segregating divided by the size of the region.
Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

@@

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), giving also $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.

After having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalised by nucleotide divergence.
        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalised by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the counfunding effect on diversity of mutation rate, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
Similarly, combined as a ratio, $\RateBetween$ depends solely on $\RateMut$
Altogether, we thus have:
\begin{gather}
    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
    \Rightarrow \frac{\RateBetween}{\RateWhithin} = 1.
\end{gather}
Altogether, $\RateBetween / \RateWhithin$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RateBetween$ and $\RateWhithin$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.

@@

\subsubsection*{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in unit of substitution per site) and the topology of the phylogenetic tree.
Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be root, in which case the distance is 0).
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

@@

Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
Their variation along the phylogeny is modelled as an $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting a the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance-rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

@@

% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
We assumed that the trait genotypic value is encoded by $\NbrLoci=5,000$ loci, with each locus contributing additively to the genotypic value.
Each effect is Normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
The phenotypic value is the sum of phenotypic value and an environmental effect, the environmental effect is normally distributed with variance $\VarEnv$.
We assumed a trait with an heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
Assuming a diploid panmictic population of size $\Ne=50$ at the root of tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generation along each branch is proportional to the branch length.

The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modelled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.

>>>>

% Including genomics, Inclusion of genomics, Linking genomics,
\title{Leveraging genomics to detect trait selection from within and between species variations}
\rhead{\scshape Trait selection from within and between species variations}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
    Conversely, using between species variation, current methods usually discard within species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

@@

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.

First, leveraging individual variations within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given locus), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
Gene expression exhibit such accumulation and divergence in expression accumulate faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\cite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Altogether, methods at the phylogenetic scale easily misclassify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimate of population size (within species) nor of time of speciation (between species).
For population geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For a phylogeneticist, our study can be seen as an extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

@@

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating divided by the size of the region.
Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the number of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

@@

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), also giving $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which new neutral alleles are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.

After having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, we define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region of interest ($d$).
This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {artworks/fig-input-output}
    \caption{
        Between species, the mean phenotypic trait value is changing along the phylogeny, allowing to estimate the between species trait variation, $\EstRateBetween$, which is normalized by nucleotide divergence.
        Within species, for each species the genetic variance allows the within species trait variation, $\EstRateWhithin$, which is  normalized by nucleotide diversity.
        $\EstNI$ is the ratio of $\EstRateBetween$ over $\EstRateWhithin$.
        Under neutral evolution, $\EstNI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\EstNI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\EstNI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow discard the confounding effect on diversity of mutation rate, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

@@

The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
Within species, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allows to compute $\pi$.
Combined as a ratio, $\RateWhithin$ depends solely on the genetic architecture of the trait as $\RateMut$.
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
Similarly, combined as a ratio, $\RateBetween$ depends solely on $\RateMut$
Altogether, we thus have:
\begin{gather}
    \RateWhithin = \RateBetween \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}}, \\
    \Rightarrow \frac{\RateBetween}{\RateWhithin} = 1.
\end{gather}
Altogether, $\RateBetween / \RateWhithin$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RateBetween$ and $\RateWhithin$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate ($\MutationRate$), nor the effective population size ($\Ne$) or time of divergence ($\Time$) need to be estimated.
Moreover, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.

@@

\subsubsection*{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the distance matrix computed from the branch lengths ($d$ as nucleotide divergence in units of substitution per site) and the topology of the phylogenetic tree.
Along the diagonal $\Distance_{\Spi,\Spi}$ are the total distances from the root of the tree to each taxon ($\Spi$).
The distance matrix is symmetric and the off-diagonal elements ($\Distance_{\Spi,\Spj}$) are total distance between the root and the common ancestor of taxa $\Spi$ and $\Spj$ (which can be the root, in which case the distance is 0).
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate~\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

@@

Even though the neutrality index can be applied for any trait, we so far assumed that several traits would to be independent of one another.
Assuming covariation between traits is also an extension that can be implemented within the multidimensional Brownian framework~\cite{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
Here we generalize to $\Ntrait$ traits evolving along the phylogeny and are correlated between them using the \textit{BayesCode} software~\cite{latrille_inferring_2021}.
Their variation along the phylogeny is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology (see section~S\ref{subsec:multivariate-brownian-process}).
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait, thus corresponding to $\NI$.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
$\Brownian$ is sampled using a Metropolis-Hastings algorithm (acceptance rejection), and the posterior distribution of $\CovarianceMatrix$ is sampled using a Gibbs sampler.
For each trait and each species, heritability ($\Heritability$) is set as a uniform distribution with user-defined boundaries.
The posterior distribution of $\EstNI$ thus allow to test for deviation from neutrality, for example by computing $\proba [\EstNI > 1 ]$ to test for evidence of diversifying selection and $\proba [\EstNI < 1 ]$ to test for evidence of stabilizing selection.

@@

% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
We assumed that the trait genotypic value is encoded by $\NbrLoci=5,000$ loci, with each locus contributing additively to the genotypic value.
Each effect is Normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and section~S\ref{subsec:genotype-phenotype-map} for theoretical formulation).
The phenotypic value is the sum of phenotypic value and an environmental effect, the environmental effect is normally distributed with variance $\VarEnv$.
We assumed a trait with a heritability of $\Heritability =0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalize the evolutionary dynamics of the system (Fig.~\ref{fig:simulator}A\&B).
Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we can simulate explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $, with mutation rate per generation $\MutationRate$.
From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from root to leaves of $0.18$ and average genetic diversity of $0.00276$.
To scale parameters with a mammalian regime, we used a mutation rate of $\MutationRate=0.00276 / 4 \Ne = 1.38e^{-5}$ per generation per locus and a total of $\Time = 0.18 / 1.38e^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch is proportional to the branch length.

The changes in log-$\MutationRate$ and log-$\Ne$ along the lineages are both modeled by a geometric Brownian process ($\brownian \left(0, \sigma_{\MutationRate}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, meaning standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$ in log-space from root to leaves.
Additionally, the instant value of log-$\Ne$ is modeled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulation starts from an initial sequence at equilibrium at the root of the tree and, at each node, the process is split until it finally reaches the leaves of the tree.
From a speciation process perspective, this is equivalent to an allopatric speciation over one generation.
